---
title: "RepData_Assignment1"
author: "ZH"
date: "Friday, March 13, 2015"
output:
  html_document:
    keep_md: yes
    toc: yes
---


```{r Loading Data & Packages,echo =TRUE,  warning=FALSE}
      library(plyr)
      library(lattice)
      options(scipen=100)     
      filename<-"C:\\Users\\zhalabi\\Documents\\RWork\\RepData_Assignment1\\activity.csv"
      Activity<-read.csv(filename)
      Activity$date <- as.Date(Activity$date)
```


```{r Steps Taken per Day, echo=TRUE}
   


      dtActivity<-subset(Activity[,1:2], is.na(Activity$steps)==F)
      dtResultbyDay<-aggregate(dtActivity[,1] , by=list(StepDate=dtActivity[,2]),sum)
      colnames(dtResultbyDay)<-c("Date", "Total") 
      plot.new()
      hist(dtResultbyDay$Total, main="Total Number of Steps Daily", breaks=10, xlab="Number of Steps")
      meanDay<-mean(dtResultbyDay$Total)
      medianDay<-median(dtResultbyDay$Total)
```

  Mean of the total number of steps taken per day: `r meanDay` 
  
  Median of the total number of steps taken per day: `r medianDay` 

```{r Average daily activity pattern, echo=TRUE}
      dtActivity<-subset(Activity[,c("steps","interval")], is.na(Activity$steps)==F)
      dtResultbyInterval<-aggregate(dtActivity[,1] , by=list(dtActivity[,2]),mean) 
      colnames(dtResultbyInterval)<-c("interval", "steps") 
      plot(dtResultbyInterval$interval,dtResultbyInterval$steps, type="l", xlab="Interval", ylab="Number of Steps",main="Average Number of Steps per Day by Interval")
      max_interval <- dtResultbyInterval[which.max(dtResultbyInterval$steps),1]
``` 

On average across all the days in the dataset, the `r max_interval`th contains the maximum number of steps


```{r Imputing missing values , echo=TRUE}
      dtActivity<-subset(Activity[,c("steps","interval")], is.na(Activity$steps)==F)
      incomplete <- sum(!complete.cases(Activity))
      dtImputed <- transform(Activity, steps = ifelse(is.na(Activity$steps), dtResultbyInterval$steps[match(Activity$interval, dtResultbyInterval$interval)], Activity$steps))
      dtImputed[as.character(dtImputed$date) == "2012-10-01", 1] <- 0
      dtStepsbyDayImputed <- aggregate(steps ~ date, dtImputed, sum)
      hist(dtStepsbyDayImputed$steps, main = paste("Total Steps Each Day"), col="blue", xlab="Number of Steps")
      hist(dtResultbyDay$Total, main = paste("Total Steps Each Day"), col="red", xlab="Number of Steps", add=T)
      legend("topright", c("Imputed", "Non-imputed"), col=c("blue", "red"), lwd=10)

      meanDayImputed <- mean(dtStepsbyDayImputed$steps)
      medianDayImputed <- median(dtStepsbyDayImputed$steps)
      meanDiff <- meanDay - meanDayImputed
      medianDiff <- medianDay - medianDayImputed
      totalDiff <- sum(dtResultbyDay$Total) - sum(dtStepsbyDayImputed$steps)
```

  The total number of missing values in the dataset: `r incomplete` 
  
  The mean of total number of steps taken per day: `r meanDay` 
  
  The median of total number of steps taken per day: `r medianDay` 
  
  The mean of total number of steps taken per day (Imputing Estimates): `r  meanDayImputed` 
  
  The median of total number of steps taken per day (Imputing Estimates): `r medianDayImputed` 
  
  The mean of the difference between real vs. estimate: `r meanDiff` 
  
  The median of the difference between real vs. estimate: `r  medianDiff` 
  
  The total of the difference between real vs. estimate: `r totalDiff` 

  Based on the summary results and the hostogram listed above, the estimated median is very close to the median of the original dataset in which the missing values were ignored. However, the mean value is differs slighltly.



```{r  differences in activity patterns between weekdays and weekends , echo=TRUE }

      names(dtImputed)<-c("steps","dt","intv")
      weekdays <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
      dtImputed$dayow = as.factor(ifelse(is.element(weekdays(as.Date(dtImputed$dt)),weekdays), "Weekday", "Weekend"))
      steps_by_interval_i <- ddply(dtImputed,.(intv,dayow),summarise,steps=mean(steps))
      xyplot(steps_by_interval_i$steps ~ steps_by_interval_i$intv|steps_by_interval_i$dayow, main="Average Steps per Day by Interval",xlab="Intervals", ylab="Steps",layout=c(1,2), type="l")
```

Generally, activities start and stop earlier during weekdays.

On Weekends, activity is less during the early morning hours but there are more activity during the day on average.  

On weekdays, activity is far less by 20:00 (8pm) but there is surge of activities after 20:00 (8pm) on weekends.


